---
title: "Prepare TCGA and CPTAC HNSCC Survival Datasets"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This notebook prepares harmonized HPV-negative HNSCC discovery and validation tables for the Python survival-modeling workflow. It builds matched RNA-seq and GISTIC/SCNA `SummarizedExperiment` objects for CPTAC and TCGA, standardizes core clinical variables, intersects shared RNA/CNV features across cohorts, and exports two wide CSV files:

- `CPTAC/data/TCGA_Discovery_Harmonized_Full_Data.csv`
- `CPTAC/data/CPTAC_Validation_Harmonized_Full_Data.csv`

These CSVs are the expected inputs for `scripts/modeling/train_survival_models.py`.

```{r}
# Data Manipulation
library(tidyverse)
library(data.table)
library(janitor)
library(here)

# Bioconductor / Genomics
library(SummarizedExperiment)
library(TCGAbiolinks)
library(Biobase)
library(S4Vectors)
library(matrixStats)
library(org.Hs.eg.db)
library(AnnotationDbi)
```

# Download CPTAC Pre-processed Data (LinkedOmics Freeze)

```{r}
# Define directories
DATA_DIR <- here("CPTAC", "data")
HNSC_DIR <- file.path(DATA_DIR, "HNSCC")

if (!dir.exists(DATA_DIR)) dir.create(DATA_DIR, recursive = TRUE)

# Download and Unzip
if (!dir.exists(HNSC_DIR)) {
  message("Downloading CPTAC HNSCC Data Freeze...")
  # FIX: Removed markdown syntax from the URL string
  zip_url <- "https://cptac-pancancer-data.s3.us-west-2.amazonaws.com/data_freeze_v1.2_reorganized/zip_files/HNSCC.zip"
  zip_file <- file.path(DATA_DIR, "HNSCC.zip")
  
  download.file(zip_url, destfile = zip_file)
  unzip(zip_file, exdir = DATA_DIR)
  file.remove(zip_file)
} else {
  message("CPTAC HNSCC Data directory already exists.")
}
```

# Load protein coding genes filter 

```{r}
# 1. Load and Prepare the Reference "Universe"
hnsc_aging_dir <- "/restricted/projectnb/agedisease/projects/oralcancer/aging/pnarnur"
autosomal_protein_genes <- readRDS(file.path(hnsc_aging_dir, "data", "raw_data", "data_summaries", "version_115_ensembl_protein_coding_autosomal_genes.RDS"))

# Ensure the reference itself is unique by Symbol to prevent exploding joins
# We prioritize the first Ensembl ID per symbol if duplicates exist in the reference
protein_genes <- autosomal_protein_genes %>%
  dplyr::filter(hgnc_symbol != "") %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
  dplyr::distinct(hgnc_symbol, .keep_all = TRUE)
```

# Create and save RNA-seq summarizedexperiment from ESet

## Combining survival and other patient metadata 

This section combines CPTAC phenotype, survival, and metadata tables, filters to HPV-negative tumors with usable overall-survival fields, and aligns patient metadata to the RNA-seq samples.

```{r}
# Load Raw Metadata Files
pheno <- fread(file.path(HNSC_DIR, "HNSCC_phenotype.txt"), data.table = FALSE)
surv  <- fread(file.path(HNSC_DIR, "HNSCC_survival.txt"),  data.table = FALSE)
meta  <- fread(file.path(HNSC_DIR, "HNSCC_meta.txt"),      data.table = FALSE)
meta  <- meta[-1, ] # Remove type description row

# Merge Clinical Data
pheno_surv <- dplyr::full_join(pheno, surv, by = c("idx" = "case_id"))
pheno_surv_meta <- dplyr::full_join(pheno_surv, meta, by = c("idx" = "case_id"))

# Load ExpressionSet specifically to get the 'pData' (Subject IDs)
eset_raw <- readRDS(file.path(DATA_DIR, "HNSC_RNAseq_eset.rds"))
pdata_raw <- pData(eset_raw) 

# Create MASTER Metadata linked to SubjectIDs
# We coalesce survival columns here once for the whole project
master_metadata <- pdata_raw %>% 
  dplyr::left_join(pheno_surv_meta, by = c("subjectID" = "idx"), suffix=c("",".y")) %>%
  dplyr::select(-ends_with(".y")) %>%
  dplyr::filter(HPV_inference == "NO") %>% # HPV-negative filter
  dplyr::mutate(OS_days = coalesce(OS_days, overall_survival), 
                OS_event = coalesce(OS_event, overall_free_status)) %>% 
  dplyr::filter(!is.na(OS_days)) %>% 
  dplyr::filter(!is.na(OS_event))
```

## Create SummarizedExperiment RNAseq

Filtering to protein coding, autosomal genes only 

```{r}
# 1. Filter Expression Data
eset <- eset_raw

# Filter Columns (Samples)
# Keep only Tumors AND samples present in our cleaned master metadata
keep_samples <- (eset$sample_type == 'tumor') & 
                (eset$subjectID %in% master_metadata$subjectID)
eset <- eset[, keep_samples]

# Filter Rows (Genes)
eset <- eset[rownames(eset) %in% protein_genes$hgnc_symbol, ] 

# This removes the "-T" suffix so it matches the metadata perfectly
colnames(eset) <- eset$subjectID

# 2. Align Metadata
# Ensure metadata rows match eset columns exactly
rnaseq_meta <- master_metadata[match(colnames(eset), master_metadata$subjectID), ]

# 3. Create SE
# This check will now PASS because both are "C3L-00977", "C3L-00987", etc.
testthat::expect_equal(colnames(eset), rnaseq_meta$subjectID)

se_RNAseq <- SummarizedExperiment(
    assays = list(DESeq2_log = exprs(eset)), 
    colData = rnaseq_meta
)

saveRDS(se_RNAseq, file.path(DATA_DIR, "CPTAC_se_RNAseq.rds"))
```

# Create GISTIC CNA summarizedexperiment 

Filtering to protein coding, autosomal genes only 

```{r}
# 1. Load and Clean GISTIC
gistic <- fread(file.path(HNSC_DIR, "HNSCC_WES_CNV_gene_gistic_level.txt"), data.table = FALSE)
gistic$clean_id <- sub("\\..*", "", gistic$idx)

# 2. Filter & Map Genes
gistic_filtered <- gistic %>%
  dplyr::inner_join(protein_genes, by = c("clean_id" = "ensembl_gene_id"))

# 3. Handle Duplicates
gistic_unique <- gistic_filtered %>%
  dplyr::distinct(hgnc_symbol, .keep_all = TRUE)

rownames(gistic_unique) <- gistic_unique$hgnc_symbol

# 4. Prepare Matrix
meta_cols <- c("idx", "clean_id", "hgnc_symbol")
count_matrix <- gistic_unique[, !colnames(gistic_unique) %in% meta_cols]

# 5. Filter Samples using MASTER metadata
# This ensures we ignore the filtering done in the RNA-seq step and look at the global valid patients
samples_to_keep <- colnames(count_matrix)[colnames(count_matrix) %in% master_metadata$subjectID]
count_matrix <- count_matrix[, samples_to_keep]

# 6. Align Metadata
gistic_meta <- master_metadata[match(colnames(count_matrix), master_metadata$subjectID), ]

# 7. Create SE
testthat::expect_equal(colnames(count_matrix), gistic_meta$subjectID)

se_gistic <- SummarizedExperiment(
  assays = list(gistic = as.matrix(count_matrix)), 
  colData = gistic_meta
)

saveRDS(se_gistic, file.path(DATA_DIR, "CPTAC_se_gistic.rds"))
```

# Harmonization & Export for SurvHive (python survival analysis libraries)

## Harmonizing the TCGA Discovery cohort data real quick 

- We will verify it has just protein coding autosomal genes (feature level filtering). Then we will verify that the metadata has two processed survival columns, OS_days (time-to-event) and OS_event (censoring status)

### TCGA RNA-seq 

```{r}
# 1. Load Data
TCGA_RNAseq <- readRDS('/restricted/projectnb/agedisease/projects/pancancer_aging_pbock/data/RNAseq/processed_SummarizedExperiments/TCGA-HNSC_RNAseq.rds')

# 2. Filter for Protein Coding Genes
# (Using the autosomal_protein_genes object loaded in previous chunks)
TCGA_RNAseq <- TCGA_RNAseq[rownames(TCGA_RNAseq) %in% autosomal_protein_genes$hgnc_symbol, ]

# 3. Process Survival Metadata
# Create unified OS_days and OS_event columns
# Note: TCGA vital_status is often "Dead"/"Alive". We convert to 1/0.
colData(TCGA_RNAseq)$OS_event <- ifelse(colData(TCGA_RNAseq)$vital_status == "Dead", 1, 0)

colData(TCGA_RNAseq)$OS_days <- ifelse(
  colData(TCGA_RNAseq)$vital_status == "Dead",
  colData(TCGA_RNAseq)$days_to_death,
  colData(TCGA_RNAseq)$days_to_last_follow_up
)

# 4. Filter Patients (HPV Negative & Valid Survival)
# We assume 'HPV.status' exists as per your snippet. Adjust levels if needed (e.g., "Negative", "Positive").
valid_samples <- colData(TCGA_RNAseq) %>%
  as.data.frame() %>%
  dplyr::filter(HPV.status == "Negative") %>%  # Keep only HPV Negative
  dplyr::filter(!is.na(OS_days) & OS_days > 0) %>% # Remove missing/zero survival
  dplyr::filter(!is.na(OS_event)) %>%
  rownames()

# Apply Filter
TCGA_RNAseq <- TCGA_RNAseq[, valid_samples]

# 5. Save Discovery RNA-seq SE
saveRDS(TCGA_RNAseq, file.path(DATA_DIR, "TCGA_se_RNAseq.rds"))
message(paste("TCGA RNA-seq processed. Remaining samples:", ncol(TCGA_RNAseq)))
```

### TCGA GISTIC 

```{r}
# 1. Load Data
root_dir <- "/restricted/projectnb/agedisease/projects/oralcancer/aging/pnarnur"
path_processed_SE <- file.path(root_dir, "data", "processed_data", "scna", "GISTIC", "SE")
TCGA_gistic <- readRDS(file.path(path_processed_SE, "TCGA-HNSC_SCNA_GISTIC_gene.RDS"))

# 2. Filter for Protein Coding Genes
TCGA_gistic <- TCGA_gistic[rownames(TCGA_gistic) %in% autosomal_protein_genes$hgnc_symbol, ]

# 3. align Samples to RNA-seq
# The TCGA barcodes usually match the first 12 or 15 characters. 
# However, if both objects come from TCGAbiolinks, the colnames are usually full barcodes.
# We intersect strictly to ensure paired data.

common_samples <- intersect(colnames(TCGA_RNAseq), colnames(TCGA_gistic))

# Subset both to the intersection (just in case GISTIC has fewer samples than RNAseq)
TCGA_RNAseq_matched <- TCGA_RNAseq[, common_samples]
TCGA_gistic_matched <- TCGA_gistic[, common_samples]

# 4. Transfer Cleaned Metadata
# We overwrite the GISTIC colData with the cleaned RNA-seq colData (containing OS_days, etc.)
# to ensure 100% metadata consistency between omics layers.
colData(TCGA_gistic_matched) <- colData(TCGA_RNAseq_matched)

#PN: additional step, let's make the five_levels assay in the GISTIC summarizedexperiment into an assay called "gistic". 
assays(TCGA_gistic_matched) <- list(gistic = assay(TCGA_gistic_matched, "five_levels"))

# 5. Save Discovery GISTIC SE
saveRDS(TCGA_gistic_matched, file.path(DATA_DIR, "TCGA_se_gistic.rds"))

# Update the RNAseq save as well if we dropped samples during intersection
saveRDS(TCGA_RNAseq_matched, file.path(DATA_DIR, "TCGA_se_RNAseq.rds"))

message(paste("TCGA GISTIC processed. Matched samples:", ncol(TCGA_gistic_matched)))
```

# EXPORT FEATURE AND SAMPLE INFO TO PYTHON, FROM BOTH DISCOVERY AND VALIDATION DATASETS 

Note: BMI is intentionally not exported because it is not available in the current TCGA mapping. It can be added later if TCGA BMI is retrieved from GDC and harmonized with CPTAC.

```{r}
# EXPORT FEATURE AND SAMPLE INFO TO PYTHON, FROM BOTH DISCOVERY AND VALIDATION DATASETS 

# BMI is intentionally omitted until it is available and harmonized in both cohorts.

# Load the SE objects
DATA_DIR <- here("CPTAC", "data")
se_rna_cptac <- readRDS(file.path(DATA_DIR, "CPTAC_se_RNAseq.rds"))
se_cnv_cptac <- readRDS(file.path(DATA_DIR, "CPTAC_se_gistic.rds"))
se_rna_tcga  <- readRDS(file.path(DATA_DIR, "TCGA_se_RNAseq.rds"))
se_cnv_tcga  <- readRDS(file.path(DATA_DIR, "TCGA_se_gistic.rds"))

# --- 0. CRITICAL: PRE-EXPORT FEATURE HARMONIZATION ---
# Find the Intersection of Genes (Features) across COHORTS
# We need genes that exist in BOTH CPTAC AND TCGA to allow for valid validation
common_rna_genes <- intersect(rownames(se_rna_cptac), rownames(se_rna_tcga))
common_cnv_genes <- intersect(rownames(se_cnv_cptac), rownames(se_cnv_tcga))

message(paste("Common RNA Genes found:", length(common_rna_genes)))
message(paste("Common CNV Genes found:", length(common_cnv_genes)))

# Subset all objects to these common features
# This ensures that RNA_GeneA in Discovery is the same column as RNA_GeneA in Validation
se_rna_tcga  <- se_rna_tcga[common_rna_genes, ]
se_cnv_tcga  <- se_cnv_tcga[common_cnv_genes, ]
se_rna_cptac <- se_rna_cptac[common_rna_genes, ]
se_cnv_cptac <- se_cnv_cptac[common_cnv_genes, ]

# --- 1. Define Feature Mapping ---
# Format: "Standard_Name" = "Original_Column_Name"

# Mapping for TCGA (Discovery) These are out clinical variables of interest
tcga_map <- c(
  "Age"             = "age_at_diagnosis",
  "Gender"          = "gender",
  "Stage"           = "ajcc_pathologic_stage",
  "T_Stage"         = "ajcc_pathologic_t",
  "N_Stage"         = "ajcc_pathologic_n",
  "Grade"           = "histological_grade",
  "Alcohol_History" = "alcohol_history",
  "Pack_Years"      = "pack_years_smoked",
  # TCGA Missing columns (will be created as NA later): BMI
  "OS_days"         = "OS_days",
  "OS_event"        = "OS_event"
)

# Mapping for CPTAC (Validation) These are out clinical variables of interest
cptac_map <- c(
  "Age"             = "age",
  "Gender"          = "gender",
  "Stage"           = "tumor_stage",
  "T_Stage"         = "patho_staging_pt",
  "N_Stage"         = "patho_staging_pn",
  "Grade"           = "histologic_grade",
  "Alcohol_History" = "alcohol_consum",
  "Pack_Years"      = "num_pack_years_sm",
  #"BMI"             = "BMI", # Unique to CPTAC We can add this in when we have it for TCGA
  "OS_days"         = "OS_days",
  "OS_event"        = "OS_event"
)

# --- 2. Processing Function ---
export_harmonized_data <- function(se_rna, se_cnv, cohort_name, col_map) {
  message(paste0("\n--- Processing ", cohort_name, " ---"))
  
  # A. Intersect Samples (Omics overlap)
  common_samples <- intersect(colnames(se_rna), colnames(se_cnv))
  se_rna_sub <- se_rna[, common_samples]
  se_cnv_sub <- se_cnv[, common_samples]
  message(paste("Matched samples:", length(common_samples)))

  # B. Extract and Rename Clinical Data
  meta_df <- as.data.frame(colData(se_rna_sub)) %>%
    tibble::rownames_to_column("patient_id")
  
  # Select only mapped columns and rename them
  # We use a loop/check to handle missing BMI in TCGA gracefully
  clin_clean <- meta_df %>% dplyr::select(patient_id)
  
  for (std_name in names(col_map)) {
    orig_name <- col_map[[std_name]]
    
    if (orig_name %in% colnames(meta_df)) {
      clin_clean[[std_name]] <- meta_df[[orig_name]]
    } else {
      # If column doesn't exist (e.g., BMI in TCGA), fill with NA
      clin_clean[[std_name]] <- NA
    }
  }
  
  # Standardize Categorical Values (Optional but recommended)
  # Unify Gender to "Male"/"Female" if capitalization differs
  if ("Gender" %in% colnames(clin_clean)) {
    clin_clean$Gender <- str_to_title(clin_clean$Gender)
  }

  # C. Extract Omics (Features as Columns)
  message("Extracting RNA matrix...")
  rna_mat <- t(assay(se_rna_sub, "DESeq2_log"))
  colnames(rna_mat) <- paste0("RNA_", colnames(rna_mat))
  
  message("Extracting CNV matrix...")
  cnv_mat <- t(assay(se_cnv_sub, "gistic"))
  colnames(cnv_mat) <- paste0("CNV_", colnames(cnv_mat))
  
  # D. Merge Clinical + Omics
  # We use cbind because we ensured order via `common_samples` earlier
  full_df <- cbind(clin_clean, as.data.frame(rna_mat), as.data.frame(cnv_mat))
  
  # E. Save
  out_file <- file.path(DATA_DIR, paste0(cohort_name, "_Harmonized_Full_Data.csv"))
  data.table::fwrite(full_df, out_file)
  message(paste("Saved:", out_file))
  message(paste("Dimensions:", nrow(full_df), "rows x", ncol(full_df), "cols"))
}

# --- 3. Execute ---
export_harmonized_data(se_rna_tcga, se_cnv_tcga, "TCGA_Discovery", tcga_map)
export_harmonized_data(se_rna_cptac, se_cnv_cptac, "CPTAC_Validation", cptac_map)
```


